---
title: pppindexr
author: Erik Ø. Sørensen
date: '2022-12-29'
slug: pppindexr
categories: []
tags: []
---

I have published an R package to github. It provides
functions to work with international comparisons
of real incomes (Geary Khamis, EKS, and CCD real
income indices). 

See the website here: https://erikosorensen.github.io/pppindexr/index.html

Here is how it is installed:

```{r message=FALSE, warning=FALSE}
# install.packages("devtools")
devtools::install_github("ErikOSorensen/pppindexr")
```

It comes with the Neary (2004) data installed, and here is how to calculate the 
three supplied indices for the 60 countries in Neary's paper, listing a random
sample of countries:

```{r}
library(pppindexr)
P <- Neary2004 |> dplyr::select(p1:p11)
Q <- Neary2004 |> dplyr::select(q1:q11)
Neary2004$eks <- eks(P,Q)
Neary2004$gk <- ygk(P,Q,
                    pop = Neary2004$pop1980, 
                    min_scale = TRUE)
Neary2004$ccd <- ccd(P,Q)
Neary2004 |> dplyr::sample_n(10) |> 
  dplyr::select(country, eks, ccd, gk) |>
  dplyr::arrange(eks) |>
  knitr::kable(digits=3) 
```


